### 2011-06-21 docs/architecture/architecture.tex
expand documentation on example code

MAESTRO models problems that are in tight hydrostatic equilibrium.
The fluid state is decomposed into a 1-d radial base state that
describes the hydrostatic structure of the star or atmosphere, and a
2- or 3-d Cartesian full state, that captures the departures from
hydrostatic equilibrium.  Two basic geometries are allowed.  A {\em
  plane-parallel} geometry assumes that the domain is thin compared to
the radius of curvature of the star, and therefore the 1-d base state
is perfectly aligned with the Cartesian state.  A {\em spherical}
geometry is for modeling an entire star.  Here, the 1-d base state is
not aligned with the Cartesian state.  Figure~\ref{fig:base_state}
shows these geometries.

\begin{figure}[tb]
\centering
\includegraphics[height=2.0in]{\archfigpath/base_grid} \hspace{0.5in}
\includegraphics[height=2.0in]{\archfigpath/base_spherical}
\caption{\label{fig:base_state} MAESTRO geometries, showing both the
  1-d base state and the full Cartesian state.  (Left) For multi-level
  problems in planar geometry, we force a direct alignment between the
  radial array cell centers and the Cartesian grid cell centers by
  allowing the radial base state spacing to change with space and
  time.  (Right) For multi-level problems in spherical geometry, since
  there is no direct alignment between the radial array cell centers
  and the Cartesian grid cell centers, we choose to fix the radial
  base state spacing across levels. Figure taken
  from~\cite{multilevel}.}
\end{figure}

@@

 Important subdirectories include:
 \begin{itemize}
 \item {\tt conductivity/}

@@

 A simple Fortran module for reading in 1-d initial model files.  This
 is used by the initialization routines to get the initial model data.

@@

   Several sub-directories containing routines to generate
   1-d initial conditions in hydrostatic equilibrium for 
   mapping onto the MAESTRO grid.  See \S~\ref{sec:initial_models}
   for details.

@@

   Each problem in MAESTRO gets it own subdirectory.  The {\tt
     GNUmakefile} in the problem directory includes the instructions
   on how to build the executable, including what modules in {\tt
     extern} are used.

@@

  The generic Makefiles that store the compilation flags for
  various platforms.  Platform/compiler-specific options are stored
  in the {\tt comps/} subdirectory.

@@

{\tt spherical\_heat} maps a spherical star (a Chandrasekhar-mass white
dwarf) onto the Cartesian domain and uses an external heat source to
cause the star to expand.  This is the 3-d analogue of the {\tt
  test\_basestate} unit test (see \S~\ref{sec:unit_tests}).  This
problem was discussed in \cite{multilevel}.

@@

To build the MAESTRO executable, simply type `{\tt make}' in the
problem directory.  The executable will have a name like {\tt
  main.Linux.Intel.exe}, where the specific parts of the name depend
on the compilers and OS used.  MAESTRO can be built with support for
parallelism (using MPI or MPI+OpenMP), or compiled as a serial code.
The top portion of the {\tt GNUmakefile} can be editted to choose the
build options.  Setting {\tt MPI := t} will build the code to use MPI
for parallelization.  Setting {\tt OMP := t} will use OpenMP.  OpenMP
is used together with MPI, with MPI distributing the grids across the
processors and within a shared-memory node, OpenMP allows many cores
to operate on the same grid.  A final build option determines whether
to build the code with support for debugging.  Setting {\tt NDEBUG :=
  t} disables debugging features.

@@

As the code runs, it will output both plotfiles and checkpoints as
well as one of more text diagnostic files (like {\tt
  maestro\_diag.out}) with integral or extrema information (like
maximum Mach number) from each timestep.  By default, the plotfiles
will be named {\tt plt}{\em nnnnn}, where the number {\em nnnnn} is
the timestep number when the file was outputted.  Similarly, the
checkpoints are named {\tt chk}{\em nnnnn}.  BoxLib plotfiles and
checkpoints are actually directories, with the data stored in
subdirectories grouped by refinement level.  Details of the simulation
(build information, number of processors used, output date, output
directory, runtime parameter values, ...) are stored in the {\tt
  job\_info} file in each plotfile directory.

@@

   \begin{center}
   \begin{tabular}{ll}
   \hline
   BC type    & integer flag \\
   \hline
   periodic             & -1 \\
   inlet (user-defined) & 11 \\
   outlet               & 12 \\
   symmetry             & 13 \\
   slip wall            & 14 \\
   no-slip wall         & 15 \\
   \hline
   \end{tabular}
   \end{center}

@@

{\tt Amrvis} is an easy-to-use visualization tool developed at LBL for
2- and 3-d datasets which can plot slices through 3-d datasets as well
as volume-renderings.  It can also very easily extract 1-d lines
through the dataset along any coordinate direction.  It is distributed
separately from the MAESTRO distribution.

@@

VisIt is a powerful, DOE-supported visualization tool for 2- and 3-d
datasets.  It can do contouring, volume rendering, streamlines, ...\
directly from BoxLib plotfiles (to open a plotfile in VisIt, point to
the {\tt Header} file inside the plotfile directory).  Details on
visit can be found at {\tt https://wci.llnl.gov/codes/visit/home.html}\,.

@@

In addition to the problems descibed above which use the full
capabilities of MAESTRO, there are a number of unit tests that
exercise only specific components of the MAESTRO solvers.  The
tests have their own drivers (a custom {\tt varden.f90}) that
initialize only the data needed for the specific test and call
specific MAESTRO routines directly.

@@

  This test initializes a 1-d radial base state quantity with a
  Gaussian distribution, maps it into the 3-d domain (assuming a
  spherical geometry) using the routines provided by the {\tt
    fill\_3d\_module} module, and then calls {\tt average} to put it
  back onto a 1-d radial array.  This way we test the accuracy of our
  procedure to map between the 1-d radial and 3-d Cartesian states.
  The output from this test was described in detail in
  \cite{multilevel}.

@@

Different MAESTRO problems are defined in subdirectories under
{\tt MAESTRO/}.  To add a problem, start by creating a new
sub-directory---this is where you will compile your problem and
store all the problem-specific files.

The minimum requirement to
define a new problem would be a {\tt GNUmakefile} which describes how
to build the application and an input file which lists the runtime
parameters.  The problem-specific executable is built in the problem
directory by typing {\tt make}.  Source files are found automatically
by searching the directories listed in the {\tt GNUmakefile}.
Customised versions of any source files placed in the problem-directory
override those with the same name found elsewhere.  Any unique
source files (and not simply a custom version of a file found
elsewhere) needs to be listed in a file call {\tt GPackage.mak}
in the problem-directory.

@@

MAESTRO models subsonic flows that are in hydrostatic equilibrium.
The solution in MAESTRO is broken up into a 1-d base state and the 2-
or 3-d full state.  The job of the 1-d base state in the algorithm is
to represent the hydrostatic structure.  The full, Cartesian state
carries the departures from hydrostatic equilibrium.  The underlying
formulation of the low Mach number equations assumes that the base
state is in hydrostatic equilibrium.  At the start of a simulation,
the initial model is read in and taken as the base state.  Therefore,
any initial model needs to already be in hydrostatic equilibrium.

The routines in {\tt MAESTRO/initial\_model/} prepare an initial model
for MAESTRO.  There are two types of routines here, the first type
take an existing 1-d initial model produced somewhere else (e.g.\ a
1-d stellar evolution code), and and map it onto a uniform grid, at
the desired resolution, using the equation of state in MAESTRO, and
using MAESTRO's discretization of hydrostatic equilibrium.  The second
type generate the initial model internally, by integrating the
condition of hydrostatic equilibrium together with a simplifying
assumption on the energy (e.g. isothermal or isentropic).  In
both cases hydrostatic equilibrium is enforced as:
\begin{equation}
\frac{p_{i+1} - p_i}{\Delta r} = \frac{1}{2} (\rho_i + \rho_{i+1})
g_{i+1/2}
\end{equation}
Here, $g_{i+1/2}$ is the edge-centered gravitational acceleration.

@@

A \layout\ is basically a \boxarray\ that knows information about other
boxes, or box ``connectivity.''  It contains additional information
that is used in filling ghost cells from other fine grids or from
coarser grids.  This information is stored as long as the layout
exists so that we don't have to recompute intersections every time we
do some operation with two multifabs that have that layout, for
example.

@@

A \multifab\ is a collection of \fab s at the same level of
refinement.  This is the primary data structure that MAESTRO
routine operate on.  A multilevel simulation stores the 
data in an array of multifabs, where the array index refers
to the refinement level.

@@

To build a \multifab, we need to provide a \layout, the number of
components to store in the multifab, and the number of ghostcells.  In
MAESTRO, the hierarchry of grids will be described by a single
\mllayout.  A \multifab\ can be declared and built at any time in a
simulation using the \mllayout, thereby allocating space at every
grid location in the simulation.  The sequence to build a \multifab\
appears as
\begin{lstlisting}[language={[95]fortran},mathescape=false]
  type(multifab) :: mfab(nlevs)
  ...
  do n = 1, nlevs
     call multifab_build(mfab(n), mla%la(n), nc, ng)
  enddo
\end{lstlisting}
Here, {\tt nc} is the number of components and {\tt ng} is the number
of ghostcells.  The \multifab\ is built one level at a time, using the
\layout\ for that level taken from the \mllayout, {\tt mla}.

@@

The fluid state (density, enthalpy, species) are stored together in a
cell-centered multi-component \multifab, typically named {\tt sold},
{\tt s1}, {\tt s2}, or {\tt snew} (depending on which time-level it
represents).  The enthalpy is stored as $(\rho h)$, and the species
are stored as partial-densities $(\rho X_k)$.

@@

@@

The {\tt average} module provides a routine {\tt average} that takes
a multilevel \multifab\ array and averages the full Cartesian data
onto the 1-d base state.

@@

   \begin{center}
   \begin{tabular}{lc}
   \hline
   key            & input quantities \\
   \hline
   eos\_input\_rt       & $\rho$, $T$ \\
   eos\_input\_rh       & $\rho$, $h$ \\
   eos\_input\_tp       & $T$, $p$ \\
   eos\_input\_rp       & $\rho$, $p$ \\
   eos\_input\_re       & $\rho$, $e$ \\
   eos\_input\_ps       & $p$, $s$ \\
   \hline
   \end{tabular}
   \end{center}

@@

The {\tt fill\_3d\_module} provides routines that map from the 1-d
base state to the full Cartesian 2- or 3-d state.  Variations in the
routines allow for cell-centered or edge-centered data on either the
base state or full Cartesian state.

@@

@@

\section{\label{sec:example} Example: Accessing State Data}

In MAESTRO, the state data is stored in a multifab array (the array
index refers to the AMR level).  A typical way to extract the state data
array looks like:

\begin{lstlisting}[language={[95]fortran},mathescape=false]
  subroutine example(s,dx,dt)

    use bl_types
    use multifab_module
    use geometry, only: dm, nlevs
    use variables, only: rho_comp
\end{lstlisting}

\noindent Here, the {\tt bl\_types} and {\tt multifab\_module} modules
bring in the basic BoxLib data types. Specifically, here, {\tt
bl\_types} defines {\tt dp\_t} which is the {\tt kind} used for
declaring double precision data, and {\tt multifab\_module} defines
the \multifab\ data type.  The {\tt geometry} module is a MAESTRO
module that provides the dimensionality ({\tt dm}) and the number of 
levels ({\tt nlevs}).  The {\tt variables} module is a MAESTRO
module that provides integer keys for indexing the state arrays.  In
this case the integer {\tt rho\_comp} refers to the location in the
state array corresponding to density.

\begin{lstlisting}[language={[95]fortran},mathescape=false]
    type(multifab) , intent(inout) :: s(:)
    real(kind=dp_t), intent(in   ) :: dx(:,:),dt
\end{lstlisting}

\noindent Next we declare the subroutine arguments.  Here, {\tt s(:)}
is our \multifab\ array that holds the state data.  The array index
in {\tt s} refers to the AMR level.

\begin{lstlisting}[language={[95]fortran},mathescape=false]
    ! Local variables
    real(kind=dp_t), pointer :: sp(:,:,:,:)
    integer :: i,n,ng_sp
    integer :: lo(dm),hi(dm)
\end{lstlisting}

\noindent Amongst the local variables we define here are a pointer,
{\tt sp}, that will point to a single \fab\ from the
\multifab\ {\tt s}.

\begin{lstlisting}[language={[95]fortran},mathescape=false]
    ng_sp = nghost(s(1))
\end{lstlisting}

\noindent Here we get the number of ghostcells for this particular
\multifab.  This will be needed to access the data stored in the
\fab s.  Note that all levels in a \multifab\ will have the same
number of ghostcells, so we can use {\tt s(1)} here.

\begin{lstlisting}[language={[95]fortran},mathescape=false]
    do n=1,nlevs
       do i = 1, s(n)%nboxes
\end{lstlisting}

\noindent To access the data, we loop over all the levels, and
all the boxes in the given level.  {\tt s(n)\%nboxes} is simply
the number of boxes in level {\tt n}.

\begin{lstlisting}[language={[95]fortran},mathescape=false]
          if ( multifab_remote(s(n), i) ) cycle
          sp => dataptr(s(n), i)
          lo =  lwb(get_box(s(n), i))
          hi =  upb(get_box(s(n), i))
\end{lstlisting}

@@

The actual data array is accessed through the {\tt dataptr} function,
which takes a \multifab\ (in this case, {\tt s(n)}) and the index of
the \boxtype\ ({\tt i}) we want.  {\tt sp} is always four-dimensional:
3 spatial dimensions and 1 for the components.

Finally, the index bounds of the box (just the data, not the ghostcells) are 
stored in the {\tt dm}-dimensional arrays {\tt lo} and {\tt hi}

\begin{lstlisting}[language={[95]fortran},mathescape=false]
          select case (dm)
          case (2)
             call example_2d(sp(:,:,1,rho_comp),ng_sp,lo,hi,dx(n,:),dt)
          case (3)
             call example_3d(sp(:,:,:,rho_comp),ng_sp,lo,hi,dx(n,:),dt)
          end select
       enddo    ! end loop over boxes

@@

  end subroutine example
\end{lstlisting}

\noindent Finally, in this example, we call either the function
{\tt example\_2d} for two-dimensional data or {\tt example\_3d}
for three-dimensional data.  Note that in the two-dimensional
case, we index the data as {\tt sp(:,:,1,rho\_comp)}.  Here a
`{\tt 1}' is used as the `z'-coordinate spatial index, since this
is a 2D problem, and the density component of the state is selected
(using the integer key {\tt rho\_comp}).

This routine will be supplimented with {\tt example\_2d} and {\tt
example\_3d}, which actually operate on the data.  The form of 
these functions is:

\begin{lstlisting}[language={[95]fortran},mathescape=false]
  subroutine example_2d(density,ng_sp,lo,hi,dx,dt)

@@

    integer        , intent(in) :: lo(:),hi(:), ng_sp
    real(kind=dp_t), intent(in) :: density(lo(1)-ng_sp:,lo(2)-ng_sp:)
    real(kind=dp_t), intent(in) :: dx(:),dt

    integer         :: i, j
    real(kind=dp_t) :: x, y

@@

          ! operate on density(i,j)
          ! ...

@@

  end subroutine example_2d
\end{lstlisting}

\noindent In this function, the bounds of the {\tt density} array
take into account the {\tt ng\_sp} ghostcells.  The {\tt j} and {\tt i}
loops loop over all the valid zones.  Coordinate information is 
computed from {\tt dx} and {\tt prob\_lo} which is the physical
lower bound of the domain.  {\tt bl\_constants\_module} declares
useful double-precision constants, like {\tt HALF} (0.5).

@@

>>>>

MAESTRO models problems that are in tight hydrostatic equilibrium.
The fluid state is decomposed into a 1D radial base state that
describes the hydrostatic structure of the star or atmosphere, and a
2- or 3D Cartesian full state, that captures the departures from
hydrostatic equilibrium.  Two basic geometries are allowed.  A {\em
  plane-parallel} geometry assumes that the domain is thin compared to
the radius of curvature of the star, and therefore the 1D base state
is perfectly aligned with the Cartesian state.  A {\em spherical}
geometry is for modeling an entire star.  Here, the 1D base state is
not aligned with the Cartesian state.  Figure~\ref{fig:base_state}
shows these geometries.

\begin{figure}[tb]
\centering
\includegraphics[height=2.0in]{\archfigpath/base_grid} \hspace{0.5in}
\includegraphics[height=2.0in]{\archfigpath/base_spherical}
\caption[MAESTRO geometries]{\label{fig:base_state} MAESTRO geometries, showing both the
  1D base state and the full Cartesian state.  (Left) For multi-level
  problems in planar geometry, we force a direct alignment between the
  radial array cell centers and the Cartesian grid cell centers by
  allowing the radial base state spacing to change with space and
  time.  (Right) For multi-level problems in spherical geometry, since
  there is no direct alignment between the radial array cell centers
  and the Cartesian grid cell centers, we choose to fix the radial
  base state spacing across levels. Figure taken
  from~\cite{multilevel}.}
\end{figure}

@@

 Important sub-directories include:
 \begin{itemize}
 \item {\tt conductivity/}

@@

 A simple Fortran module for reading in 1D initial model files.  This
 is used by the initialization routines to get the initial model data.

@@

   Several sub-directories containing routines to generate
   1D initial conditions in hydrostatic equilibrium for 
   mapping onto the MAESTRO grid.  See \S~\ref{sec:initial_models}
   for details.

@@

   Each problem in MAESTRO gets it own sub-directory.  The {\tt
     GNUmakefile} in the problem directory includes the instructions
   on how to build the executable, including what modules in {\tt
     extern} are used.

@@

  The generic Makefiles that store the compilation flags for
  various platforms.  Platform/compiler-specific options are stored
  in the {\tt comps/} sub-directory.

@@

{\tt spherical\_heat} maps a spherical star (a Chandrasekhar-mass white
dwarf) onto the Cartesian domain and uses an external heat source to
cause the star to expand.  This is the 3D analogue of the {\tt
  test\_basestate} unit test (see \S~\ref{sec:unit_tests}).  This
problem was discussed in \cite{multilevel}.

@@

To build the MAESTRO executable, simply type `{\tt make}' in the
problem directory.  The executable will have a name like {\tt
  main.Linux.Intel.exe}, where the specific parts of the name depend
on the compilers and OS used.  MAESTRO can be built with support for
parallelism (using MPI or MPI+OpenMP), or compiled as a serial code.
The top portion of the {\tt GNUmakefile} can be edited to choose the
build options.  Setting {\tt MPI := t} will build the code to use MPI
for parallelization.  Setting {\tt OMP := t} will use OpenMP.  OpenMP
is used together with MPI, with MPI distributing the grids across the
processors and within a shared-memory node, OpenMP allows many cores
to operate on the same grid.  A final build option determines whether
to build the code with support for debugging.  Setting {\tt NDEBUG :=
  t} disables debugging features.

@@

As the code runs, it will output both plotfiles and checkpoints as
well as one of more text diagnostic files (like {\tt
  maestro\_diag.out}) with integral or extrema information (like
maximum Mach number) from each timestep.  By default, the plotfiles
will be named {\tt plt}{\em nnnnn}, where the number {\em nnnnn} is
the timestep number when the file was outputted.  Similarly, the
checkpoints are named {\tt chk}{\em nnnnn}.  BoxLib plotfiles and
checkpoints are actually directories, with the data stored in
sub-directories grouped by refinement level.  Details of the simulation
(build information, number of processors used, output date, output
directory, runtime parameter values, ...) are stored in the {\tt
  job\_info} file in each plotfile directory.

@@

   \begin{center}
   \begin{tabular}{ll}
   \hline
   BC type    & integer flag \\
   \hline
   periodic             & $-1$ \\
   inlet (user-defined) & $11$ \\
   outlet               & $12$ \\
   symmetry             & $13$ \\
   slip wall            & $14$ \\
   no-slip wall         & $15$ \\
   \hline
   \end{tabular}
   \end{center}

@@

{\tt Amrvis} is an easy-to-use visualization tool developed at LBL for
2- and 3D datasets which can plot slices through 3D datasets as well
as volume-renderings.  It can also very easily extract 1D lines
through the dataset along any coordinate direction.  It is distributed
separately from the MAESTRO distribution.

@@

VisIt is a powerful, DOE-supported visualization tool for 2- and 3D
datasets.  It can do contouring, volume rendering, streamlines, ...\
directly from BoxLib plotfiles (to open a plotfile in VisIt, point to
the {\tt Header} file inside the plotfile directory).  Details on
visit can be found at {\tt https://wci.llnl.gov/codes/visit/home.html}\,.

@@

In addition to the problems described above which use the full
capabilities of MAESTRO, there are a number of unit tests that
exercise only specific components of the MAESTRO solvers.  The
tests have their own drivers (a custom {\tt varden.f90}) that
initialize only the data needed for the specific test and call
specific MAESTRO routines directly.

@@

  This test initializes a 1D radial base state quantity with a
  Gaussian distribution, maps it into the 3D domain (assuming a
  spherical geometry) using the routines provided by the {\tt
    fill\_3d\_module} module, and then calls {\tt average} to put it
  back onto a 1D radial array.  This way we test the accuracy of our
  procedure to map between the 1D radial and 3D Cartesian states.
  The output from this test was described in detail in
  \cite{multilevel}.

@@

Different MAESTRO problems are defined in sub-directories under
{\tt MAESTRO/}.  To add a problem, start by creating a new
sub-directory---this is where you will compile your problem and
store all the problem-specific files.

The minimum requirement to
define a new problem would be a {\tt GNUmakefile} which describes how
to build the application and an input file which lists the runtime
parameters.  The problem-specific executable is built in the problem
directory by typing {\tt make}.  Source files are found automatically
by searching the directories listed in the {\tt GNUmakefile}.
Customized versions of any source files placed in the problem-directory
override those with the same name found elsewhere.  Any unique
source files (and not simply a custom version of a file found
elsewhere) needs to be listed in a file call {\tt GPackage.mak}
in the problem-directory.

@@

MAESTRO models subsonic flows that are in hydrostatic equilibrium.
The solution in MAESTRO is broken up into a 1D base state and the 2-
or 3D full state.  The job of the 1D base state in the algorithm is
to represent the hydrostatic structure.  The full, Cartesian state
carries the departures from hydrostatic equilibrium.  The underlying
formulation of the low Mach number equations assumes that the base
state is in hydrostatic equilibrium.  At the start of a simulation,
the initial model is read in and taken as the base state.  Therefore,
any initial model needs to already be in hydrostatic equilibrium.

The routines in {\tt MAESTRO/initial\_model/} prepare an initial model
for MAESTRO.  There are two types of routines here, the first type
take an existing 1D initial model produced somewhere else (e.g.\ a
1D stellar evolution code), and and map it onto a uniform grid, at
the desired resolution, using the equation of state in MAESTRO, and
using MAESTRO's discretization of hydrostatic equilibrium.  The second
type generate the initial model internally, by integrating the
condition of hydrostatic equilibrium together with a simplifying
assumption on the energy (e.g. isothermal or isentropic).  In
both cases hydrostatic equilibrium is enforced as:
\begin{equation}
\frac{p_{i+1} - p_i}{\Delta r} = \frac{1}{2} (\rho_i + \rho_{i+1})
g_{i+1/2}
\end{equation}
Here, $g_{i+1/2}$ is the edge-centered gravitational acceleration.

@@

A \layout\ is basically a \boxarray\ that knows information about other
boxes, or box ``connectivity.''  It contains additional information
that is used in filling ghost cells from other fine grids or from
coarser grids.  This information is stored as long as the layout
exists so that we don't have to recompute intersections every time we
do some operation with two \multifab s that have that layout, for
example.

@@

A \multifab\ is a collection of \fab s at the same level of
refinement.  This is the primary data structure that MAESTRO
routine operate on.  A multilevel simulation stores the 
data in an array of \multifab s, where the array index refers
to the refinement level.

@@

To build a \multifab, we need to provide a \layout, the number of
components to store in the \multifab\, and the number of ghostcells.  In
MAESTRO, the hierarchy of grids will be described by a single
\mllayout.  A \multifab\ can be declared and built at any time in a
simulation using the \mllayout, thereby allocating space at every
grid location in the simulation.  The sequence to build a \multifab\
appears as
\begin{lstlisting}[language={[95]fortran},mathescape=false]
  type(multifab) :: mfab(nlevs)
  ...
  do n = 1, nlevs
     call multifab_build(mfab(n), mla%la(n), nc, ng)
  enddo
\end{lstlisting}
Here, {\tt nc} is the number of components and {\tt ng} is the number
of ghostcells.  The \multifab\ is built one level at a time, using the
\layout\ for that level taken from the \mllayout, {\tt mla}.

@@

The fluid state (density, enthalpy, species, temperature, and tracer)
are stored together in a cell-centered multi-component \multifab,
typically named {\tt sold}, {\tt s1}, {\tt s2}, or {\tt snew}
(depending on which time-level it represents).  The enthalpy is stored
as $(\rho h)$, and the species are stored as partial-densities $(\rho
X_k)$.  The tracer component is not used at present time, but can
describe an arbitrary advected quantity.

@@

Note: the pressure is not carried as part of the `{\tt s}' \multifab s.

@@

The {\tt average} module provides a routine {\tt average} that takes
a multilevel \multifab\ array and averages the full Cartesian data
onto the 1D base state.

@@

   \begin{center}
   \begin{tabular}{lc}
   \hline
   key            & input quantities \\
   \hline
   {\tt eos\_input\_rt}       & $\rho$, $T$ \\
   {\tt eos\_input\_rh}       & $\rho$, $h$ \\
   {\tt eos\_input\_tp}       & $T$, $p$ \\
   {\tt eos\_input\_rp}       & $\rho$, $p$ \\
   {\tt eos\_input\_re}       & $\rho$, $e$ \\
   {\tt eos\_input\_ps}       & $p$, $s$ \\
   \hline
   \end{tabular}
   \end{center}

@@

The {\tt fill\_3d\_module} provides routines that map from the 1D
base state to the full Cartesian 2- or 3D state.  Variations in the
routines allow for cell-centered or edge-centered data on either the
base state or full Cartesian state.

@@

Other keys in the {\tt variables} modules are reserved for boundary
conditions ({\tt foextrap\_comp} and {\tt hoextap\_comp}), the
projection of the pressure ({\tt press\_comp}), or constructing
the plotfile.

@@

\section{\label{sec:example} Example: Accessing State and MAC Data}

In MAESTRO, the state data is stored in a cell-centered multifab array
(the array index refers to the AMR level) and the MAC velocities are
stored in a 2D nodal multifab array (with indices referring to the AMR
level and the velocity component).  Here we demonstrate a typical way
to extract the state and MAC velocity data.

All MAESTRO routines are contained in a module, to allow for compile-time
argument checking.
\begin{lstlisting}[language={[95]fortran},mathescape=false]
module example_module

contains
\end{lstlisting}

The main interface to our routine is called {\tt example}---this will
take the \multifab s containing the data and then pass them to the
work routines, {\tt example\_2d} or {\tt example\_3d}, depending on
the dimensionality.  
\begin{lstlisting}[language={[95]fortran},mathescape=false]
  subroutine example(mla,s,umac,dx,dt)

    use bl_types
    use multifab_module
    use ml_layout_module
    use variables, only: rho_comp
\end{lstlisting}

\noindent Here, the {\tt bl\_types} and {\tt multifab\_module} modules
bring in the basic BoxLib data types. Specifically, here, {\tt
  bl\_types} defines {\tt dp\_t} which is the {\tt kind} used for
declaring double precision data, and {\tt multifab\_module} defines
the \multifab\ data type.  The {\tt ml\_layout\_module} defines the
datatype for a \mllayout---many routines will take an \mllayout
to allow us to fill ghostcells.  The {\tt variables} module is a MAESTRO
module that provides integer keys for indexing the state arrays.  In
this case the integer {\tt rho\_comp} refers to the location in the
state array corresponding to density.

Next we declare the subroutine arguments:

\begin{lstlisting}[language={[95]fortran},mathescape=false]
    type(ml_layout), intent(in   ) :: mla
    type(multifab) , intent(inout) :: s(:)
    type(multifab) , intent(inout) :: umac(:,:)
    real(kind=dp_t), intent(in   ) :: dx(:,:),dt
\end{lstlisting}
Here, {\tt s(:)} is our \multifab\ array that holds the state data.
with the array index in {\tt s} refers to the AMR level.  The MAC 
velocities are held in the multifab {\tt umac}, with the array
indices referring to the AMR level and the component.

Local variable declarations come next:
\begin{lstlisting}[language={[95]fortran},mathescape=false]
    ! Local variables
    real(kind=dp_t), pointer :: sp(:,:,:,:)
    real(kind=dp_t), pointer :: ump(:,:,:,:), vmp(:,:,:,:), wmp(:,:,:,:)
    integer :: i,n,dm,nlevs,ng_sp,ng_um
    integer :: lo(mla%dim),hi(mla%dim)
\end{lstlisting}

\noindent Amongst the local variables we define here are a pointer,
{\tt sp}, that will point to a single \fab\ from the
\multifab\ {\tt s}, and a pointer for each component of the MAC
velocity, {\tt ump}, {\tt vmp}, and {\tt wmp} (for a 2D run,
we won't use {\tt wmp}).  We note that regardless of the dimensionality,
these pointers are 4-dimensional: 3 spatial + 1 component.

Next we get the dimensionality and number of levels
\begin{lstlisting}[language={[95]fortran},mathescape=false]
    dm = mla%dim
    nlevs = mla%nlevel
\end{lstlisting}

Each multifab can have their own number of ghostcells, so we get
these next:
\begin{lstlisting}[language={[95]fortran},mathescape=false]
    ng_sp = nghost(s(1))
    ng_um = nghost(umac(1,1))
\end{lstlisting}
By convention, all levels in a given multifab have the same number of
ghostcells, so we use level 1 in the {\tt nghost()} call.  We also use
the same number of ghostcells for each component of the velocity, so
we only need to consider the first component in the {\tt nghost()}
call.  The ghostcells will be needed to access the data stored in the
\fab s.

To access the data, we loop over all the levels, and all the boxes in
the given level.
\begin{lstlisting}[language={[95]fortran},mathescape=false]
    do n=1,nlevs
       do i = 1, nboxes(s(n))
\end{lstlisting}
{\tt nboxes(s(n))} is simply the number of boxes in level {\tt n}.
All data in all \multifab s will have the same box layout, so we only
need to query the number of boxes from one \multifab.

For a given \boxtype, we get the data and the bounds of the \boxtype.
\begin{lstlisting}[language={[95]fortran},mathescape=false]
          if ( multifab_remote(s(n), i) ) cycle
          sp  => dataptr(s(n), i)
          ump => dataptr(umac(n,1),i)
          vmp => dataptr(umac(n,2),i)
          lo =  lwb(get_box(s(n), i))
          hi =  upb(get_box(s(n), i))
\end{lstlisting}

@@

The actual data array is accessed through the {\tt dataptr} function,
which takes a \multifab\ (e.g.\ {\tt s(n)}) and the index of the
\boxtype\ ({\tt i}) we want.  We see that the $x$ MAC velocity for the
current box is stored in {\tt ump} and the $y$ MAC velocity is stored
in {\tt vmp}.  We don't get the $z$ velocity data here, since that
would not be available for a 2D run---we defer that until we test on
the dimensionality below.

Finally, the index bounds of the box (just the data, not the ghostcells) are 
stored in the {\tt dm}-dimensional arrays {\tt lo} and {\tt hi}.  These indices
refer to the current box, and hold for both the state, {\tt sp}, and the MAC
velocity, {\tt ump} and {\tt vmp}.  However, since the MAC velocity is nodal
in the component direction, the loops over the valid data will differ
slight (as we see below).

With the data extracted, we call a subroutine to operate on it.  We use
different subroutines for the different dimensionalities (and many times
have a separate routine for spherical geometries).
\begin{lstlisting}[language={[95]fortran},mathescape=false]
          select case (dm)
          case (2)
             call example_2d(sp(:,:,1,rho_comp),ng_sp, &
                             ump(:,:,1,1),vmp(:,:,1,1),ng_um, &
                             lo,hi,dx(n,:),dt)
          case (3)
             wmp => dataptr(umac(n,3),i)
             call example_3d(sp(:,:,:,rho_comp),ng_sp, &
                             ump(:,:,:,1),vmp(:,:,:,1),wmp(:,:,:,1),ng_um, &
                             lo,hi,dx(n,:),dt)
          end select
       enddo    ! end loop over boxes

@@

  end subroutine example
\end{lstlisting}
\noindent We call either the function
{\tt example\_2d} for two-dimensional data or {\tt example\_3d}
for three-dimensional data.  Note that in the two-dimensional
case, we index the data as {\tt sp(:,:,1,rho\_comp)}.  Here a
`{\tt 1}' is used as the `z'-coordinate spatial index, since this
is a 2D problem, and the density component of the state is selected
(using the integer key {\tt rho\_comp}).  The 3D version accesses
the data as {\tt sp(:,:,:,rho\_comp)}---only the component regarding
the variable is needed here.  Notice that we also pass through
the number of ghostcells for each of the quantities.

This routine will be supplimented with {\tt example\_2d} and {\tt
example\_3d}, which actually operate on the data.  The form of 
the 2D function is:

\begin{lstlisting}[language={[95]fortran},mathescape=false]
  subroutine example_2d(density,ng_sp, &
                        umac,vmac,ng_um, &
                        lo,hi,dx,dt)

@@

    integer        , intent(in) :: lo(:),hi(:), ng_sp, ng_um
    real(kind=dp_t), intent(in) :: density(lo(1)-ng_sp:,lo(2)-ng_sp:)
    real(kind=dp_t), intent(in) ::    umac(lo(1)-ng_um:,lo(2)-ng_um:)
    real(kind=dp_t), intent(in) ::    vmac(lo(1)-ng_um:,lo(2)-ng_um:)

    real(kind=dp_t), intent(in) :: dx(:),dt

    integer         :: i, j
    real(kind=dp_t) :: x, y
    real(kind=dp_t) :: dens, u, v

@@

          dens = density(i,j)

          ! compute cell-centered velocity
          u = HALF*(umac(i,j) + umac(i+1,j))
          v = HALF*(vmac(i,j) + vmac(i+1,j))

          ! operate on the data
          ! ...

@@

  end subroutine example_2d

end module example_module
\end{lstlisting}

\noindent In this function, the bounds of the {\tt density} array take
into account the {\tt ng\_sp} ghostcells and the index space of the
current box.  Likewise, the MAC velocities refer to the {\tt ng\_um}
ghostcells.  The {\tt j} and {\tt i} loops loop over all the valid
zones.  Coordinate information is computed from {\tt dx} and {\tt
  prob\_lo} which is the physical lower bound of the domain.  {\tt
  bl\_constants\_module} declares useful double-precision constants,
like {\tt HALF} (0.5).  Here, we see how to access the density for
the current zone and compute the cell-centered velocities from the
MAC velocities.  By convection, for a nodal array, the indices refer 
to the {\em lower} interface in the nodal direction, so for {\tt umac},
{\tt umac(i,j)} and {\tt umac(i+1,j)} are the $x$ MAC velocities
on the lower and upper edge of the zone in the $x$-direction.

@@

In this example, we looped over the valid zones.  If we wished to loop
over the interfaces bounding the valid zones, in the $x$-direction,
we would loop as
\begin{lstlisting}[language={[95]fortran},mathescape=false]
  do j = lo(2), hi(2)
     do i = lo(1), hi(1)+1
        ! access umac(i,j)
     enddo
  enddo
\end{lstlisting}
